---
title: 'Visualise PDTO kinase screening data'
author: 
  - name: 'Momeneh (Sepideh) Foroutan'
    affiliation: 'Prof Hollande Lab'
    url: https://www.monash.edu/discovery-institute/huntington-lab
date: '2019-06-10'
output:
  rmdformats::readthedown:
    fig_width: 12
    fig_height: 6
    gallery: TRUE
    highlight: tango
    lightbox: TRUE
    self_contained: TRUE
    thumbnails: FALSE
    number_sections: TRUE	
    toc_depth: 4
    use_bookdown: TRUE
    code_folding: hide
  html_document2:
    df_print: paged
params:
  update_date: !r paste("Last updated on:", Sys.Date() )
editor_options: 
  chunk_output_type: inline
---

`r params$update_date`

<style>
body {
text-align: justify}
</style>


# Set up and overview
This document contains code for reproducing the results of the drug screening data on PDTOs, reported in [Behrenbruch and Foroutan et al manuscript](https://www.biorxiv.org/content/10.1101/2021.02.04.429849v1).
We generate heatmaps in Figure 6a.
```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

options(max.print = "75")

opts_chunk$set(
  cache = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  root.dir = normalizePath("."))

opts_knit$set(width = 65)
```

```{r}
library(RColorBrewer)
library(ComplexHeatmap)
library(ggplot2)

figPath <- "../figure/Screening/"
dataPath <- "../data/Screening/"
outPath <- "../output/Screening/"
```

```{r}
dd <- read.csv(paste0(dataPath, "CTG_OnlyLowSeedingFiltered_Pathways_Targets-FH.csv"), stringsAsFactors = F)

dd$Dose <- sapply(dd$Treatment, function(x){
  xx <- strsplit(x, "_")
  xx[[1]][length(xx[[1]])]
})

dd$Group_CTG <- "No hit"
dd$Group_CTG[dd$Bin == "LC"] <- "Hit"

dd$Drug <- dd$Treatment
dd$Drug <- gsub("_5$", "", dd$Drug)
dd$Drug <- gsub("_0.5$", "", dd$Drug)
dd$Drug <- gsub("_0$", "", dd$Drug)
dd$Drug <- gsub("_25$", "", dd$Drug)
dd$Drug <- gsub("_250$", "", dd$Drug)
dd$Drug <- gsub("_0.25$", "", dd$Drug)
dd$Drug <- gsub("_2.5$", "", dd$Drug)
dd$Drug <- gsub("_0.25_0.025$", "", dd$Drug)

cc <- c("Cell-cycle progression","cell-cycle progression", "cell-cycle progression ", "Cell-cycle progression ")

dd$Pathway[dd$Pathway  %in% cc] <- "Cell cycle progression"

dd$Pathway <- gsub(" Pathway", "", dd$Pathway)
dd$Pathway <- gsub(" PATHWAY", "", dd$Pathway)
dd$Pathway <- gsub(" pathway", "", dd$Pathway)
dd$Pathway <- gsub(" Signaling", "", dd$Pathway)
dd$Pathway <- gsub(" Signalling", "", dd$Pathway)
dd$Pathway <- gsub(" signalling", "", dd$Pathway)
dd$Pathway <- gsub("TGF beta", "TGFb", dd$Pathway)
dd$Pathway <- gsub("JNK and p38 MAPK", "JNK/p38/MAPK", dd$Pathway)
dd$Pathway <- gsub("NK-Kappa B", "NFkB", dd$Pathway)
dd$Pathway <- gsub("miscellaneous", "Miscellaneous", dd$Pathway)

dd$Target <- gsub("miscellaneous", "Miscellaneous", dd$Target)


##----- replace everything above 100 with 100 in percFC.
dd$Perc_FC[dd$Perc_FC > 100] <- 100
dd$ZScore[dd$ZScore > 15] <- 15
dd$ZScore[dd$ZScore < -15] <- -15

colnames(dd)[colnames(dd) == "Group_CTG"] <- "Hit"
dd$Hit[dd$Hit == "Hit"] <- "Yes"
dd$Hit[dd$Hit == "No hit"] <- "No"

```


```{r}
# pdf("DensityPlot_PercFC_Zscore.pdf", height = 5, width = 6)
# ggplot(dd, aes(x = Perc_FC, color = Patient)) +
#   geom_density(size = 1) + theme_bw()
# 
# ggplot(dd, aes(x = ZScore, color = Patient)) +
#   geom_density(size = 1) + theme_bw()
# dev.off()
```


# Heatmap for each PDTO
Read the data containing the Mahalanobis values.
## PDTO1: yPDB516_LT1
```{r}
currentPat <- "yPDB516_LT1"

## 624
dd1 <- dd[dd$Patient == currentPat, ]
dd1_hits <- dd1[dd1$Hit == "Yes", ]

## 858
m1 <- read.csv(paste0(dataPath, "CB_PMC15_PrimaryScreen_mp_P01.csv"), 
  stringsAsFactors = F, check.names = F)

m1$treatment.id <- gsub("uM", "", m1$treatment.id)
m1$Mahalanobis[m1$Mahalanobis > 150] <- 150
# m1$Mahalanobis[m1$Mahalanobis < 5] <- 5

## 624 common
length(intersect(m1$treatment.id, dd1$Treatment))


m1_hits <- m1[m1$mp.value < 0.05 & m1$Mahalanobis > 5, ]
# m1_hits <- m1[m1$mp.value < 0.05 , ]


## 87 common (87 and 185)
length(intersect(dd1_hits$Treatment, m1_hits$treatment.id))


dd1 <- merge(dd1, m1, by.x = "Treatment", by.y = "treatment.id")
```

Comment/Uncomment the section below to generate plots for different doses (0.5 vs 5).
```{r}

# ##---- for dose 0.5
# dd1_0.5 <- dd1[dd1$Dose == "0.5", ]
# ##-- For ordering based on Pathway
# dd1_0.5 <- dplyr::arrange(dd1_0.5, Pathway, Perc_FC, ZScore, Mahalanobis)
# ##-- for ordering based on Target
# rownames(dd1_0.5) <- dd1_0.5$Treatment
# pData1 <- dd1_0.5
# currentDose <- "0.5"


##---- for dose 5
dd1_5 <- dd1[dd1$Dose == "5", ]
##-- For ordering based on Pathway
dd1_5 <- dplyr::arrange(dd1_5, Pathway, Perc_FC, ZScore, Mahalanobis)
##-- for ordering based on Target
# dd1_5 <- dplyr::arrange(dd1_5, Target, ZScore, Perc_FC)
rownames(dd1_5) <- dd1_5$Treatment
pData1 <- dd1_5
currentDose <- "5"

```

```{r}
##------------ set parametres

textSize <- 18
# annotColumn <- "Target"
annotColumn <- "Pathway"


pathwayCols <- c(brewer.pal(8, "Set1"),
  brewer.pal(8, "Dark2")[-c(5, 7)],
  brewer.pal(8, "Accent")[1:2]
  )

names(pathwayCols) <- unique(pData1[, annotColumn])

# groupCol <- c("gray20", "gray80")
# groupCol <- brewer.pal(8, "Dark2")[1:2]

##---------  for patient 1
pathwayHm1 <- Heatmap(pData1[, annotColumn, drop = F],
  name = annotColumn,
  col = pathwayCols,
  width = unit(5, "mm") , 
  row_order = rownames(pData1), 
  show_row_names = F,
  show_heatmap_legend = T,
  column_names_gp = gpar(fontsize = textSize),
   heatmap_legend_param = list(
          title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
          labels_gp = gpar(fontsize = textSize),
          grid_height = unit(0.8, "cm")
  ))


HitCol <- c("gray10", "gray90")
names(HitCol) <- c("Yes", "No")

HitHm1 <- Heatmap(pData1[, "Hit", drop = F],
  name = "\nHit",
  col = HitCol,
  width = unit(5, "mm") , 
  row_order = rownames(pData1), 
  show_row_names = F,
  show_heatmap_legend = T,
  column_names_gp = gpar(fontsize = textSize),
   heatmap_legend_param = list(
          title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
          labels_gp = gpar(fontsize = textSize),
          grid_height = unit(0.8, "cm")
  ))

ctgHm1 <- Heatmap(
  pData1[, grepl("ZScore", colnames(pData1)), drop = F],
  # col = viridis::viridis(100), 
  col = circlize::colorRamp2(c(
    -15,
    0,
    15
  ),
    c("navy",
      "azure",
      "yellow3")),
  width = unit(12, "mm"),
  show_row_names = F,
  cluster_rows = F,
  column_title = "yP516_LT1",
  column_title_gp = gpar(fontsize = textSize + 2),
  column_names_gp = gpar(fontsize = textSize),
  # row_order = rownames(dd1),
  # name = paste0("CTG Z-score\n",
  #   "[",
  #   round(min(pData1$ZScore), 1),
  #   ", ",
  #   round(max(pData1$ZScore), 1), "]"),
  heatmap_legend_param = list(
    title = "\nCTG Z-score\n",
    title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
    labels_gp = gpar(fontsize = textSize),
    color_bar = "continuous", at = c(-15, 0, 15),  legend_height = unit(3, "cm")
  )
)

fcHm1 <- Heatmap(pData1[, grepl("Perc_FC", colnames(pData1)), drop = F], 
  col = viridis::magma(100),
  width = unit(12, "mm"), 
  cluster_rows = F, 
  show_row_names = F,
  column_names_gp = gpar(fontsize = textSize),
  # row_names_gp = gpar(fontsize = 8),
  # name = paste0("Percent FC\n", 
  #   "[", 
  #   round(min(pData1$Perc_FC), 1), 
  #   ", ", 
  #   round(max(pData1$Perc_FC), 1), "]"),
  heatmap_legend_param = list(
          title = "\nPercent FC\n",
          title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
          labels_gp = gpar(fontsize = textSize),
          color_bar = "continuous",  at =c(0, 50, 100), legend_height = unit(3, "cm")
        ))

mahHm1 <- Heatmap(
  pData1[, grepl("Mahalanobis", colnames(pData1)), drop = F],
  # col = viridis::viridis(100), 
  col = rev(viridis::viridis(100, end = 0.90)),
  width = unit(12, "mm"),
  show_row_names = F,
  cluster_rows = F,
  # column_title = "yP516_LT1",
  column_title_gp = gpar(fontsize = textSize + 2),
  column_names_gp = gpar(fontsize = textSize),
  heatmap_legend_param = list(
    title = "\nMahalanobis\n",
    title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
    labels_gp = gpar(fontsize = textSize),
    color_bar = "continuous",
    at = c(0, 75, 150),
    legend_height = unit(3, "cm")
  )
)


pdf(
  paste0(figPath,
    "Figure_6a_CTG_yP516_LT1_Dose",
    currentDose,
    "_",
    annotColumn,
    ".pdf"
  ),
  height = 14,
  width = 10
)
draw(
   pathwayHm1 +  fcHm1 + ctgHm1 + mahHm1 + HitHm1 ,
  heatmap_legend_side = "left",
      legend_title_gp = gpar(fontsize = textSize, fontface = "italic"),
  ht_gap = unit(c(2, 2, 2, 2), "mm")
    )
dev.off()
```

## PDTO2: yPJF520_LT1
```{r}
currentPat <- "yPJF520_LT1"

## 624
dd2 <- dd[dd$Patient == currentPat, ]
dd2_hits <- dd2[dd2$Hit == "Yes", ]

## 858
m2 <- read.csv(paste0(dataPath, "CB_PMC15_PrimaryScreen_mp_P02.csv"), 
  stringsAsFactors = F, check.names = F)

m2$treatment.id <- gsub("uM", "", m2$treatment.id)
m2$Mahalanobis[m2$Mahalanobis > 150] <- 150

## 624 common
length(intersect(m2$treatment.id, dd2$Treatment))


m2_hits <- m2[m2$mp.value < 0.05 & m2$Mahalanobis > 5, ]
# m1_hits <- m1[m1$mp.value < 0.05 , ]

## 94 common (124 and 198) --> 76% are confirmed
length(intersect(dd2_hits$Treatment, m2_hits$treatment.id))

dd2 <- merge(dd2, m2, by.x = "Treatment", by.y = "treatment.id")

```


Comment/Uncomment the section below to generate plots for different doses (0.5 vs 5).
```{r}
#---- for dose 0.5
dd1_0.5 <- dd2[dd2$Dose == "0.5", ]
##-- For ordering based on Pathway
dd1_0.5 <- dplyr::arrange(dd1_0.5, Pathway, Perc_FC, ZScore)
##-- for ordering based on Target
# dd1_0.5 <- dplyr::arrange(dd1_0.5, Target, ZScore, Perc_FC)
rownames(dd1_0.5) <- dd1_0.5$Treatment
pData1 <- dd1_0.5
currentDose <- "0.5"


# ##---- for dose 5
# dd1_5 <- dd2[dd2$Dose == "5", ]
# ##-- For ordering based on Pathway
# dd1_5 <- dplyr::arrange(dd1_5, Pathway, Perc_FC, ZScore)
# ##-- for ordering based on Target
# # dd1_5 <- dplyr::arrange(dd1_5, Target, ZScore, Perc_FC)
# rownames(dd1_5) <- dd1_5$Treatment
# pData1 <- dd1_5
# currentDose <- "5"
```

```{r}
##------------ set parametres

textSize <- 18
# annotColumn <- "Target"
annotColumn <- "Pathway"


pathwayCols <- c(brewer.pal(8, "Set1"),
  brewer.pal(8, "Dark2")[-c(5, 7)],
  brewer.pal(8, "Accent")[1:2]
  )

names(pathwayCols) <- unique(pData1[, annotColumn])


##---------  for patient 1
pathwayHm1 <- Heatmap(pData1[, annotColumn, drop = F],
  name = annotColumn,
  col = pathwayCols,
  width = unit(5, "mm") , 
  row_order = rownames(pData1), 
  show_row_names = F,
  show_heatmap_legend = T,
  column_names_gp = gpar(fontsize = textSize),
   heatmap_legend_param = list(
          title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
          labels_gp = gpar(fontsize = textSize),
          grid_height = unit(0.8, "cm")
  ))


HitCol <- c("gray10", "gray90")
names(HitCol) <- c("Yes", "No")

HitHm1 <- Heatmap(pData1[, "Hit", drop = F],
  name = "\nHit",
  col = HitCol,
  width = unit(5, "mm") , 
  row_order = rownames(pData1), 
  show_row_names = F,
  show_heatmap_legend = T,
  column_names_gp = gpar(fontsize = textSize),
   heatmap_legend_param = list(
          title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
          labels_gp = gpar(fontsize = textSize),
          grid_height = unit(0.8, "cm")
  ))

ctgHm1 <- Heatmap(
  pData1[, grepl("ZScore", colnames(pData1)), drop = F],
  # col = viridis::viridis(100), 
  col = circlize::colorRamp2(c(
    -15,
    0,
    15
  ),
    c("navy",
      "azure",
      "yellow3")),
  width = unit(12, "mm"),
  show_row_names = F,
  cluster_rows = F,
  column_title = "yP520_LT1",
  column_title_gp = gpar(fontsize = textSize + 2),
  column_names_gp = gpar(fontsize = textSize),
  heatmap_legend_param = list(
    title = "\nCTG Z-score\n",
    title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
    labels_gp = gpar(fontsize = textSize),
    color_bar = "continuous", at = c(-15, 0, 15),  legend_height = unit(3, "cm")
  )
)

fcHm1 <- Heatmap(pData1[, grepl("Perc_FC", colnames(pData1)), drop = F], 
  col = viridis::magma(100),
  width = unit(12, "mm"), 
  cluster_rows = F, 
  show_row_names = F,
  column_names_gp = gpar(fontsize = textSize),
  heatmap_legend_param = list(
          title = "\nPercent FC\n",
          title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
          labels_gp = gpar(fontsize = textSize),
          color_bar = "continuous",  at = c(0, 50, 100), legend_height = unit(3, "cm")
        ))


mahHm1 <- Heatmap(
  pData1[, grepl("Mahalanobis", colnames(pData1)), drop = F],
  col = rev(viridis::viridis(100, end = 0.90)),
  width = unit(12, "mm"),
  show_row_names = F,
  cluster_rows = F,
  column_title_gp = gpar(fontsize = textSize + 2),
  column_names_gp = gpar(fontsize = textSize),
  heatmap_legend_param = list(
    title = "\nMahalanobis\n",
    title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
    labels_gp = gpar(fontsize = textSize),
    color_bar = "continuous",
    at = c(0, 75, 150),
    legend_height = unit(3, "cm")
  )
)

pdf(
  paste0(figPath,
    "Figure_6a_CTG_yP520_LT1_Dose",
    currentDose,
    "_",
    annotColumn,
    ".pdf"
  ),
  height = 14,
  width = 10
)
draw(
   pathwayHm1 +  fcHm1 + ctgHm1 + mahHm1 +HitHm1 ,
  heatmap_legend_side = "left",
      legend_title_gp = gpar(fontsize = textSize, fontface = "italic"),
  ht_gap = unit(c(2, 2, 2, 2), "mm")
    )
dev.off()
```


## PDTO3: yPMK295_LT
```{r}
currentPat <- "yPMK295_LT"

## 624
dd3 <- dd[dd$Patient == currentPat, ]
dd3_hits <- dd3[dd3$Hit == "Yes", ]

## 858
m3 <- read.csv(paste0(dataPath, "CB_PMC15_PrimaryScreen_mp_P03.csv"), 
  stringsAsFactors = F, check.names = F)

m3$treatment.id <- gsub("uM", "", m3$treatment.id)
m3$Mahalanobis[m3$Mahalanobis > 150] <- 150


m3_hits <- m3[m3$mp.value < 0.05 & m3$Mahalanobis > 5, ]

## 88 common (90 and 208): 98%
length(intersect(dd3_hits$Treatment, m3_hits$treatment.id))


m2_hits <- m2[m2$mp.value < 0.05 & m2$Mahalanobis > 5, ]

dd3 <- merge(dd3, m3, by.x = "Treatment", by.y = "treatment.id")
```


Comment/Uncomment the section below to generate plots for different doses (0.5 vs 5).
```{r}
##---- for dose 0.5
dd1_0.5 <- dd3[dd3$Dose == "0.5", ]
##-- For ordering based on Pathway
dd1_0.5 <- dplyr::arrange(dd1_0.5, Pathway, Perc_FC, ZScore)
##-- for ordering based on Target
# dd1_0.5 <- dplyr::arrange(dd1_0.5, Target, ZScore, Perc_FC)
rownames(dd1_0.5) <- dd1_0.5$Treatment
pData1 <- dd1_0.5
currentDose <- "0.5"


##---- for dose 5
# dd1_5 <- dd3[dd3$Dose == "5", ]
# ##-- For ordering based on Pathway
# dd1_5 <- dplyr::arrange(dd1_5, Pathway, Perc_FC, ZScore)
# ##-- for ordering based on Target
# # dd1_5 <- dplyr::arrange(dd1_5, Target, ZScore, Perc_FC)
# rownames(dd1_5) <- dd1_5$Treatment
# pData1 <- dd1_5
# currentDose <- "5"
```

```{r}
##------------ set parametres

textSize <- 18
annotColumn <- "Pathway"


pathwayCols <- c(brewer.pal(8, "Set1"),
  brewer.pal(8, "Dark2")[-c(5, 7)],
  brewer.pal(8, "Accent")[1:2]
  )

names(pathwayCols) <- unique(pData1[, annotColumn])


##---------  for patient 1
pathwayHm1 <- Heatmap(pData1[, annotColumn, drop = F],
  name = annotColumn,
  col = pathwayCols,
  width = unit(5, "mm") , 
  row_order = rownames(pData1), 
  show_row_names = F,
  show_heatmap_legend = T,
  column_names_gp = gpar(fontsize = textSize),
   heatmap_legend_param = list(
          title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
          labels_gp = gpar(fontsize = textSize),
          grid_height = unit(0.8, "cm")
  ))


HitCol <- c("gray10", "gray90")
names(HitCol) <- c("Yes", "No")

HitHm1 <- Heatmap(pData1[, "Hit", drop = F],
  name = "\nHit",
  col = HitCol,
  width = unit(5, "mm") , 
  row_order = rownames(pData1), 
  show_row_names = F,
  show_heatmap_legend = T,
  column_names_gp = gpar(fontsize = textSize),
   heatmap_legend_param = list(
          title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
          labels_gp = gpar(fontsize = textSize),
          grid_height = unit(0.8, "cm")
  ))

ctgHm1 <- Heatmap(
  pData1[, grepl("ZScore", colnames(pData1)), drop = F],
  # col = viridis::viridis(100), 
  col = circlize::colorRamp2(c(
    -15,
    0,
    15
  ),
    c("navy",
      "azure",
      "yellow3")),
  width = unit(12, "mm"),
  show_row_names = F,
  cluster_rows = F,
  column_title = "yP295_LT1",
  column_title_gp = gpar(fontsize = textSize + 2),
  column_names_gp = gpar(fontsize = textSize),
  heatmap_legend_param = list(
    title = "\nCTG Z-score\n",
    title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
    labels_gp = gpar(fontsize = textSize),
    color_bar = "continuous", at = c(-15, 0, 15),  legend_height = unit(3, "cm")
  )
)

fcHm1 <- Heatmap(pData1[, grepl("Perc_FC", colnames(pData1)), drop = F], 
  col = viridis::magma(100),
  width = unit(12, "mm"), 
  cluster_rows = F, 
  show_row_names = F,
  column_names_gp = gpar(fontsize = textSize),
  heatmap_legend_param = list(
          title = "\nPercent FC\n",
          title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
          labels_gp = gpar(fontsize = textSize),
          color_bar = "continuous",  at =c(0, 50, 100), legend_height = unit(3, "cm")
        ))


mahHm1 <- Heatmap(
  pData1[, grepl("Mahalanobis", colnames(pData1)), drop = F],
  # col = viridis::viridis(100), 
  col = rev(viridis::viridis(100, end = 0.90)),
  width = unit(12, "mm"),
  show_row_names = F,
  cluster_rows = F,
  # column_title = "yP516_LT1",
  column_title_gp = gpar(fontsize = textSize + 2),
  column_names_gp = gpar(fontsize = textSize),
  heatmap_legend_param = list(
    title = "\nMahalanobis\n",
    title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
    labels_gp = gpar(fontsize = textSize),
    color_bar = "continuous",
    at = c(0, 75, 150),
    legend_height = unit(3, "cm")
  )
)

pdf(
  paste0(figPath,
    "Figure_6a_CTG_yP295_LT_Dose",
    currentDose,
    "_",
    annotColumn,
    ".pdf"
  ),
  height = 14,
  width = 10
)
draw(
   pathwayHm1 +  fcHm1 + ctgHm1 + mahHm1 + HitHm1 ,
  heatmap_legend_side = "left",
      legend_title_gp = gpar(fontsize = textSize, fontface = "italic"),
  ht_gap = unit(c(2, 2, 2, 2), "mm")
    )
dev.off()
```


### combine data deom all patients
```{r}
dds <- Reduce(rbind, list(dd1, dd2, dd3))

dds$Patient <- gsub("yPMK295", "yP295", dds$Patient)
dds$Patient <- gsub("yPDB516", "yP516", dds$Patient)
dds$Patient <- gsub("yPJF520", "yP520", dds$Patient)

dds <- dds[, !grepl("original", colnames(dds))]

write.csv(dds, paste0(outPath, "CTG_FC_Mahalanobis_3patients.csv"), row.names = F)
```


# Heatmap for all PDTOs 
Generate heatmaps by including data from all PDTOs to summarise teh results of the primary screening. Some of these results reproduces the two heatmaps represented in Suppl Figure 11a.
```{r}
dd$Patient <- gsub("yPMK295", "yP295", dd$Patient)
dd$Patient <- gsub("yPDB516", "yP516", dd$Patient)
dd$Patient <- gsub("yPJF520", "yP520", dd$Patient)
```

## Dose 0.5
### CTG
```{r}
library(tidyr)

textSize <- 15

dd0 <- dd %>% 
  dplyr::select(Patient, Treatment, ZScore, Pathway, Dose) %>% 
  dplyr::filter(Dose == "0.5") %>% 
  tidyr::pivot_wider(names_from = Patient, values_from = ZScore) %>% 
  data.frame()

annotColumn <- "Pathway"
pathwayCols <- c(brewer.pal(8, "Set1"),
  brewer.pal(8, "Dark2")[-c(5, 7)],
  brewer.pal(8, "Accent")[1:2]
  )

names(pathwayCols) <- unique(dd0[, annotColumn])


pathwayHm0 <- HeatmapAnnotation(Pathway = dd0[, annotColumn],  
  which = "row", show_annotation_name = F,
  # name = "",
  col = list(Pathway = pathwayCols),
  width = unit(5, "mm") ,
   annotation_legend_param = list(
          title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
          labels_gp = gpar(fontsize = textSize),
     grid_height = unit(0.8, "cm")
  )
)

ctgHm0 <- Heatmap(
  dd0[, grepl("^yP", colnames(dd0)), drop = F],
  col = circlize::colorRamp2(c(-15,
    0,
    15),
    c("navy",
      "azure",
      "yellow3")),
  width = unit(30, "mm"),
  show_row_names = F,
  cluster_rows = T,
  column_title = "CTG\n(0.5)",
  
  column_title_gp = gpar(fontsize = textSize + 2),
  column_names_gp = gpar(fontsize = textSize),
  name = paste0("CTG Z-score\n",
    "[",
    round(min(dd$ZScore), 1),
    ", ",
    round(max(dd$ZScore), 1), "]"),
  heatmap_legend_param = list(
    title = "\nCTG Z-score\n",
    title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
    labels_gp = gpar(fontsize = textSize),
    color_bar = "continuous",
    at = c(-15, 0, 15),
    legend_height = unit(3, "cm")
  ),
  left_annotation = pathwayHm0
)

pdf(
  paste0(figPath, "CTG_3Patients_Dose0.5_",  annotColumn, ".pdf"),
  height = 10,
  width = 6
)
draw(ctgHm0 , 
  heatmap_legend_side = "left",
      legend_title_gp = gpar(fontsize = textSize, fontface = "italic"),
  ht_gap = unit(c(8), "mm"))
dev.off()
```

### FC
```{r}

library(tidyr)
dd00 <- dd %>%
  dplyr::select(Patient, Treatment, Perc_FC, Pathway, Dose) %>%
  dplyr::filter(Dose == 0.5) %>%
  tidyr::pivot_wider(names_from = Patient, values_from = Perc_FC) %>%
  data.frame()

annotColumn <- "Pathway"
pathwayCols <- c(brewer.pal(8, "Set1"),
  brewer.pal(8, "Dark2")[-c(5, 7)],
  brewer.pal(8, "Accent")[1:2]
  )

names(pathwayCols) <- unique(dd00[, annotColumn])


pathwayHm00 <- HeatmapAnnotation(Pathway = dd00[, annotColumn],  
  which = "row", show_annotation_name = F,
  col = list(Pathway = pathwayCols),
  width = unit(5, "mm") ,
   annotation_legend_param = list(
          title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
          labels_gp = gpar(fontsize = textSize),
     grid_height = unit(0.8, "cm")
  )
)


pats <- c("yP516_LT1", "yP520_LT1", "yP295_LT")


ctgHm00 <- Heatmap(
  dd00[, pats, drop = F],
  col = viridis::magma(100),
  width = unit(30, "mm"),
  show_row_names = F,
  cluster_rows = T,
  cluster_columns = F,
  column_title = "FC\n(0.5)",
  name = paste0("Percent FC\n",
    "[",
    round(min(dd$Perc_FC), 1),
    ", ",
    round(max(dd$Perc_FC), 1), "]"),
  heatmap_legend_param = list(
    title = "\nPercent FC\n",
    title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
    labels_gp = gpar(fontsize = textSize),
    color_bar = "continuous", at = c(0, 50, 100), legend_height = unit(3, "cm")
  ), left_annotation = pathwayHm00
)


pdf(
  paste0(figPath, "Supp_Figure_11a_PercFC_3Patients_Dose0.5_",  annotColumn, ".pdf"),
  height = 10,
  width = 6
)
draw(ctgHm00 , 
  heatmap_legend_side = "left",
      legend_title_gp = gpar(fontsize = textSize, fontface = "italic"),
  ht_gap = unit(c(8), "mm"))
dev.off()
# 
```

### Mahalanobis
```{r}

getCols <- c("Treatment", "Mahalanobis", "Pathway", "Dose")

ddM <-
  merge(dd1[, getCols], dd2[, c("Treatment", "Mahalanobis")], by = "Treatment")
colnames(ddM)[colnames(ddM) == "Mahalanobis.x"] <- "yP516_LT1"
colnames(ddM)[colnames(ddM) == "Mahalanobis.y"] <- "yP520_LT1"

ddM <-
  merge(ddM, dd3[, c("Treatment", "Mahalanobis")], by = "Treatment")
colnames(ddM)[colnames(ddM) == "Mahalanobis"] <- "yP295_LT"



library(tidyr)

dd000 <- ddM %>%
  dplyr::filter(Dose == 0.5) %>%
  data.frame()

annotColumn <- "Pathway"
pathwayCols <- c(brewer.pal(8, "Set1"),
  brewer.pal(8, "Dark2")[-c(5, 7)],
  brewer.pal(8, "Accent")[1:2]
  )

names(pathwayCols) <- unique(dd[, annotColumn])

pathwayHm00 <- HeatmapAnnotation(Pathway = dd000[, annotColumn],  
  which = "row", show_annotation_name = F,
  col = list(Pathway = pathwayCols),
  width = unit(5, "mm") ,
   annotation_legend_param = list(
          title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
          labels_gp = gpar(fontsize = textSize),
     grid_height = unit(0.8, "cm")
  )
)


pats <- c("yP516_LT1", "yP520_LT1", "yP295_LT")


mahHm00 <- Heatmap(
  dd000[, pats, drop = F],
  col = rev(viridis::viridis(100, end = 0.90)),
  width = unit(30, "mm"),
  show_row_names = F,
  cluster_rows = T,
  cluster_columns = F,
  column_title = "Mahalanobis\n(0.5)",
  column_title_gp = gpar(fontsize = textSize + 2),
  column_names_gp = gpar(fontsize = textSize),
  heatmap_legend_param = list(
    title = "\nMahalanobis\n",
    title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
    labels_gp = gpar(fontsize = textSize),
    color_bar = "continuous",
    at = c(0, 75, 150),
    legend_height = unit(3, "cm")
  ), left_annotation = pathwayHm00
)

pdf(
  paste0(figPath, "Mahalanobis_3Patients_Dose0.5_",  annotColumn, ".pdf"),
  height = 10,
  width = 6
)
draw(mahHm00 , 
  heatmap_legend_side = "left",
      legend_title_gp = gpar(fontsize = textSize, fontface = "italic"),
  ht_gap = unit(c(8), "mm"))
dev.off()
```




## Dose 5
### CTG
```{r}
library(tidyr)

textSize <- 15

dd0 <- dd %>% 
  dplyr::select(Patient, Treatment, ZScore, Pathway, Dose) %>% 
  dplyr::filter(Dose == "5") %>% 
  tidyr::pivot_wider(names_from = Patient, values_from = ZScore) %>% 
  data.frame()

annotColumn <- "Pathway"
pathwayCols <- c(brewer.pal(8, "Set1"),
  brewer.pal(8, "Dark2")[-c(5, 7)],
  brewer.pal(8, "Accent")[1:2]
  )

names(pathwayCols) <- unique(dd0[, annotColumn])

pathwayHm0 <- HeatmapAnnotation(Pathway = dd0[, annotColumn],  
  which = "row", show_annotation_name = F,
  col = list(Pathway = pathwayCols),
  width = unit(5, "mm") ,
   annotation_legend_param = list(
          title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
          labels_gp = gpar(fontsize = textSize),
     grid_height = unit(0.8, "cm")
  )
)

ctgHm0 <- Heatmap(
  dd0[, grepl("^yP", colnames(dd0)), drop = F],
  col = circlize::colorRamp2(c(-15,
    0,
    15),
    c("navy",
      "azure",
      "yellow3")),
  width = unit(30, "mm"),
  show_row_names = F,
  cluster_rows = T,
  column_title = "CTG\n(5)",
  
  column_title_gp = gpar(fontsize = textSize + 2),
  column_names_gp = gpar(fontsize = textSize),
  name = paste0("CTG Z-score\n",
    "[",
    round(min(dd$ZScore), 1),
    ", ",
    round(max(dd$ZScore), 1), "]"),
  heatmap_legend_param = list(
    title = "\nCTG Z-score\n",
    title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
    labels_gp = gpar(fontsize = textSize),
    color_bar = "continuous",
    at = c(-15, 0, 15),
    legend_height = unit(3, "cm")
  ),
  left_annotation = pathwayHm0
)

pdf(
  paste0(figPath, "CTG_3Patients_Dose5_",  annotColumn, ".pdf"),
  height = 10,
  width = 6
)
draw(ctgHm0 , 
  heatmap_legend_side = "left",
      legend_title_gp = gpar(fontsize = textSize, fontface = "italic"),
  ht_gap = unit(c(8), "mm"))
dev.off()
```

### FC
```{r}

library(tidyr)
dd00 <- dd %>%
  dplyr::select(Patient, Treatment, Perc_FC, Pathway, Dose) %>%
  dplyr::filter(Dose == 5) %>%
  tidyr::pivot_wider(names_from = Patient, values_from = Perc_FC) %>%
  data.frame()

annotColumn <- "Pathway"
pathwayCols <- c(brewer.pal(8, "Set1"),
  brewer.pal(8, "Dark2")[-c(5, 7)],
  brewer.pal(8, "Accent")[1:2]
  )

names(pathwayCols) <- unique(dd00[, annotColumn])

pathwayHm00 <- HeatmapAnnotation(Pathway = dd00[, annotColumn],  
  which = "row", show_annotation_name = F,
  col = list(Pathway = pathwayCols),
  width = unit(5, "mm") ,
   annotation_legend_param = list(
          title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
          labels_gp = gpar(fontsize = textSize),
     grid_height = unit(0.8, "cm")
  )
)


pats <- c("yP516_LT1", "yP520_LT1", "yP295_LT")


ctgHm00 <- Heatmap(
  dd00[, pats, drop = F],
  col = viridis::magma(100),
  width = unit(30, "mm"),
  show_row_names = F,
  cluster_rows = T,
  cluster_columns = F,
  column_title = "FC\n(5)",
  name = paste0("Percent FC\n",
    "[",
    round(min(dd$Perc_FC), 1),
    ", ",
    round(max(dd$Perc_FC), 1), "]"),
  heatmap_legend_param = list(
    title = "\nPercent FC\n",
    title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
    labels_gp = gpar(fontsize = textSize),
    color_bar = "continuous", at = c(0, 50, 100), legend_height = unit(3, "cm")
  ), left_annotation = pathwayHm00
)



pdf(
  paste0(figPath, "Supp_Figure_11a_PercFC_3Patients_Dose5_",  annotColumn, ".pdf"),
  height = 10,
  width = 6
)
draw(ctgHm00 , 
  heatmap_legend_side = "left",
      legend_title_gp = gpar(fontsize = textSize, fontface = "italic"),
  ht_gap = unit(c(8), "mm"))
dev.off()
# 
```

### Mahalanobis
```{r}

getCols <- c("Treatment", "Mahalanobis", "Pathway", "Dose")

ddM <-
  merge(dd1[, getCols], dd2[, c("Treatment", "Mahalanobis")], by = "Treatment")
colnames(ddM)[colnames(ddM) == "Mahalanobis.x"] <- "yP516_LT1"
colnames(ddM)[colnames(ddM) == "Mahalanobis.y"] <- "yP520_LT1"

ddM <-
  merge(ddM, dd3[, c("Treatment", "Mahalanobis")], by = "Treatment")
colnames(ddM)[colnames(ddM) == "Mahalanobis"] <- "yP295_LT"



library(tidyr)

dd000 <- ddM %>%
  dplyr::filter(Dose == 5) %>%
  data.frame()

annotColumn <- "Pathway"
pathwayCols <- c(brewer.pal(8, "Set1"),
  brewer.pal(8, "Dark2")[-c(5, 7)],
  brewer.pal(8, "Accent")[1:2]
  )

names(pathwayCols) <- unique(dd[, annotColumn])

pathwayHm00 <- HeatmapAnnotation(Pathway = dd000[, annotColumn],  
  which = "row", show_annotation_name = F,
  col = list(Pathway = pathwayCols),
  width = unit(5, "mm") ,
   annotation_legend_param = list(
          title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
          labels_gp = gpar(fontsize = textSize),
     grid_height = unit(0.8, "cm")
  )
)


pats <- c("yP516_LT1", "yP520_LT1", "yP295_LT")


mahHm00 <- Heatmap(
  dd000[, pats, drop = F],
  col = rev(viridis::viridis(100, end = 0.90)),
  width = unit(30, "mm"),
  show_row_names = F,
  cluster_rows = T,
  cluster_columns = F,
  column_title = "Mahalanobis\n(5)",
  column_title_gp = gpar(fontsize = textSize + 2),
  column_names_gp = gpar(fontsize = textSize),
  heatmap_legend_param = list(
    title = "\nMahalanobis\n",
    title_gp = gpar(fontsize = textSize + 2, fontface = "italic"),
    labels_gp = gpar(fontsize = textSize),
    color_bar = "continuous",
    at = c(0, 75, 150),
    legend_height = unit(3, "cm")
  ), left_annotation = pathwayHm00
)

pdf(
  paste0(figPath, "Mahalanobis_3Patients_Dose5_",  annotColumn, ".pdf"),
  height = 10,
  width = 6
)
draw(mahHm00 , 
  heatmap_legend_side = "left",
      legend_title_gp = gpar(fontsize = textSize, fontface = "italic"),
  ht_gap = unit(c(8), "mm"))
dev.off()
```


